
```{r, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Metacell Analysis Toolkit (MATK) {#command-line}

We provide a command line tool allowing users to build metacells using either tool (MC2, SuperCell or SEACells) from a provided dataset.
The command line tool takes multiple parameters as input, *e.g.,* number of neighbors considered in the knn, number of components used, graining level.
which is for example required in a benchmark setting.

We will use this toolkit to analyse differentiating hematopoietic stem cells (Human CD34+).
Note that we retrieved the RNA data from a 10X multiome experiment but we won't analyse the ATAC modality. 
To retrieve this data, please refer to section \@ref(CD34-data).


## Setting up the environment
Please skip this step if you did not use conda in the requirements section.
```{r, eval = FALSE}
library(reticulate)
conda_env <-  conda_list()[reticulate::conda_list()$name == "MetacellAnalysisToolkit","python"]

Sys.setenv(RETICULATE_PYTHON = conda_env)
```


```{r}
library(SuperCell)
library(Seurat)
# If you have Seurat V5 installed, specify that you want to analyze Seurat V4 objects
if(packageVersion("Seurat") >= 5) {options(Seurat.object.assay.version = "v4"); print("you are using seurat v5 with assay option v4")}
library(anndata)
library(MetacellAnalysisToolkit)
library(ggplot2)
```
## Metacell building{#MC-continuous}
Here we will use the MATK bash command line to build the metacells with MetaCell2 and SuperCell. 
To call the MATK command line, please define your path to the gihub cloned repository optained from this [github repository](https://github.com/GfellerLab/MetacellToolkit).

### Building metacell with MetaCell2 (MC2)
We use MetaCell2 (`-t MetaCell`) to identify metacells at a gamma of 50 (`-g 50`). We specify that we want a Seurat .rds object in ouput (`-s seurat`).
```{bash, eval = T, echo = T}
#git clone https://github.com/GfellerLab/MetacellAnalysisToolkit
MATK_path=MetacellAnalysisToolkit/
${MATK_path}/cli/MATK -t MetaCell -i data/CD34/cd34_multiome_rna.h5ad -o data/CD34/MC2/ -g 50 -s seurat
```

### Building metacell with SuperCell
We use SuperCell (`-t SuperCell`) to identify metacells at a gamma of 50 (`-g 50`). We specify that we want a Seurat .rds object in ouput (`-s seurat`).
We use 2000 highly variable genes (HVGs, `-f 2000`) to compute a PCA from which we use 50 components (`-n 50`) to make a k = 30 knn (`-k 30`) graph on which we identify the metacells.
```{bash, eval = T, echo = T}
MATK_path=MetacellAnalysisToolkit/
${MATK_path}/cli/MATK -t SuperCell -i data/CD34/cd34_multiome_rna.h5ad -o data/CD34/SuperCell/ -f 2000 -g 50 -n 50 -k 30 -s seurat
```

## Short downstream analysis of the metacells

### MetaCell2 metacells
We load the object obtained with MATK
```{r}
cd34.metacell <- readRDS("data/CD34/MC2/mc_Seurat.rds")
cd34.metacell
```
We Normalize the metacells data, identify the HVGs, and we make a classical dimensionality reduction (first a PCA and then vizualisation with a UMAP).
```{r}
cd34.metacell <- NormalizeData(cd34.metacell)
cd34.metacell <- FindVariableFeatures(cd34.metacell)
cd34.metacell <- ScaleData(cd34.metacell)
cd34.metacell <- RunPCA(cd34.metacell)
cd34.metacell <- RunUMAP(cd34.metacell,dims = c(1:50), min.dist = 0.5)
```

Plot the results using Seurat.
```{r}
umap.metacell <- UMAPPlot(cd34.metacell,group.by = "celltype",label = T)
umap.metacell
```

We correctly retrieve the two main differentiation path of CD34+ cells.
In one hand, Megakaryocyte Erythrocyte Progenitor (MEP) fate with GATA2 and then GATA1 transcription factor expression.
On the other, Lymphoid Myeloid Multipotent Progenitor (LMPP) fate with SPI1 transcription factor expression.

```{r}
genes <- c("CD34","GATA2","GATA1","SPI1")
marker.metacell.umaps <- lapply(genes, FUN = function(x){
  plot <- FeaturePlot(cd34.metacell, features = x)
  }
)
marker.metacell.umaps
```
When analyzing metacells it's a good idea to plot their size on these. We can do it with ggplot.
```{r}
data <- cbind(umap.metacell$data, cd34.metacell$size)
colnames(data)[1:2] <- c("umap_1", "umap_2")
colnames(data)[length(colnames(data))] <- "size"
ggplot(data,aes(x = umap_1, y = umap_2, color = celltype)) + geom_point(aes(size=size)) + theme_classic()
```
```{r}
lapply(marker.metacell.umaps,FUN = function(x){
  data <- cbind(x$data,cd34.metacell$size)
  colnames(data)[1:2] <- c("umap_1", "umap_2")
  colnames(data)[length(colnames(data))] <- "size"
  gene <- colnames(data)[length(colnames(data))-1]
  ggplot(data,aes(x = umap_1, y = umap_2, size=size, colour = .data[[gene]])) + geom_point() + 
    scale_color_gradientn(colors =  c("lightgrey", "blue"), 
                          guide = "colorbar")  +
  theme_classic() + ggtitle(gene)
} )

```

### SuperCell metacells

We can do the same with the metacells obtained with SuperCell

We load the object obtained with MATK
```{r}
cd34.supercell <- readRDS("data/CD34/SuperCell/mc_Seurat.rds")
cd34.supercell
```
We Normalize the supercells data, identify the HVGs, and we make a classical dimensionality reduction (first a PCA and then vizualisation with a UMAP).
```{r}
cd34.supercell <- NormalizeData(cd34.supercell)
cd34.supercell <- FindVariableFeatures(cd34.supercell)
cd34.supercell <- ScaleData(cd34.supercell)
cd34.supercell <- RunPCA(cd34.supercell)
cd34.supercell <- RunUMAP(cd34.supercell,dims = c(1:50),min.dist = 0.5)
```

Plot the results using Seurat.
```{r}
umap.supercell <- UMAPPlot(cd34.supercell,group.by = "celltype",label = T)
umap.supercell
```
We correctly retrieve the two main differentiation path of CD34+ cells.
In one hand, Megakaryocyte Erythrocyte Progenitor (MEP) fate with GATA2 and then GATA1 transcription factor expression.
On the other, Lymphoid Myeloid Multipotent Progenitor (LMPP) fate with SPI1 transcription factor expression.
```{r}
genes <- c("CD34","GATA2","GATA1","SPI1")
marker.supercell.umaps <- lapply(genes, FUN = function(x){
  plot <- FeaturePlot(cd34.supercell, features = x)
  }
)
marker.supercell.umaps
```
When analyzing supercells it's a good idea to plot their size on these. We can do it with ggplot.
```{r}
data <- cbind(umap.supercell$data,cd34.supercell$size)
colnames(data)[1:2] <- c("umap_1", "umap_2")
colnames(data)[length(colnames(data))] <- "size"
ggplot(data,aes(x = umap_1, y = umap_2, color = celltype)) + geom_point(aes(size=size)) + theme_classic()
```

```{r}
lapply(marker.supercell.umaps,FUN = function(x){
  data <- cbind(x$data,cd34.supercell$size)
  colnames(data)[1:2] <- c("umap_1", "umap_2")
  colnames(data)[length(colnames(data))] <- "size"
  gene <- colnames(data)[length(colnames(data))-1]
  ggplot(data,aes(x = umap_1, y = umap_2, size=size,colour = .data[[gene]])) + geom_point() + 
    scale_color_gradientn(colors =  c("lightgrey", "blue"), 
                          guide = "colorbar")  +
  theme_classic() + ggtitle(gene)
} )

```

## QC with the MetacellAnalysisToolkit package
### Loading single-cell data

First we need to load the single cell data and make a classical dimentionality reduction analysis.
```{r}
cd34.singlecells <- read_h5ad("data/CD34/cd34_multiome_rna.h5ad")
cd34.singlecells <- CreateSeuratObject(counts = Matrix::t(cd34.singlecells$X),meta.data = cd34.singlecells$obs)
cd34.singlecells <- NormalizeData(cd34.singlecells)
if(packageVersion("Seurat") >= 5) {
  cd34.singlecells[["RNA"]] <- as(object = cd34.singlecells[["RNA"]], Class = "Assay")
}
cd34.singlecells <- FindVariableFeatures(cd34.singlecells)
cd34.singlecells <- ScaleData(cd34.singlecells)
cd34.singlecells <- RunPCA(cd34.singlecells)
cd34.singlecells <- RunUMAP(cd34.singlecells,dims = c(1:50))
```
Plot single cell data.
```{r}
UMAPPlot(cd34.singlecells,group.by = "celltype",label = T)
```
### Visualization of metacells in single-cell space

Now we can plot metacells from MetaCell2 in the single-cell space.
```{r}
mc_projection(
  sc.obj = cd34.singlecells,
  mc.obj = cd34.metacell,
  cell.membership = cd34.metacell@misc$membership,
  sc.reduction = "umap",
  sc.label = "celltype", # single cells will be colored according the sc.label
  metacell.label = "celltype" # metacells cell will be colored according the metacell.label
  ) + theme_classic()
```

And same for metacells obtained with SuperCell.
```{r}
mc_projection(
  sc.obj = cd34.singlecells,
  mc.obj = cd34.supercell,
  cell.membership = cd34.supercell@misc$cell_membership,
  sc.reduction = "umap",
  sc.label = "celltype", # single cells will be colored according the sc.label
  metacell.label = "celltype" # metacells cell will be colored according the metacell.label
  ) + theme_classic()
```
### Compactness and separation

We can compute the compactness and separation of the metacells from the single cell pca. 
We can use the approach of SEACells by computing these metrics in a diffusion map obtained from the pca using the `get_diffusion_comp` function and then
run the `mc_compactnness` and `mc_separation` functions to obtain the compactness and separation metrics.

First for MetaCell2 metacells
```{r}
diffusion_comp <- get_diffusion_comp(sc.obj = cd34.singlecells, sc.reduction = "pca", dims = 1:50)
cd34.metacell$compactness <- mc_compactness(cell.membership = cd34.metacell@misc$cell_membership, 
                                            sc.obj = cd34.singlecells,
                                            group.label = "metacell_name",
                                            sc.reduction = diffusion_comp,
                                            dims = 1:ncol(diffusion_comp))

qc_boxplot(mc.obj = cd34.metacell, qc.metrics = "compactness")
```

```{r}
cd34.metacell$separation <- mc_separation(cell.membership = cd34.metacell@misc$cell_membership,
                                          sc.obj = cd34.singlecells,
                                          group.label = "metacell_name",
                                          sc.reduction = diffusion_comp,
                                          dims =  1:ncol(diffusion_comp))

qc_boxplot(mc.obj = cd34.metacell, qc.metrics = "separation")

```
Same for metacell obtained with SuperCell.
```{r}
cd34.supercell$compactness <- mc_compactness(cell.membership = cd34.supercell@misc$cell_membership, 
                                            sc.obj = cd34.singlecells,
                                            sc.reduction = diffusion_comp, 
                                            dims = 1:ncol(diffusion_comp))

qc_boxplot(mc.obj = cd34.supercell, qc.metrics = "compactness")
```

```{r}
cd34.supercell$separation <- mc_separation(cell.membership = cd34.supercell@misc$cell_membership, 
                                          sc.obj = cd34.singlecells,
                                          sc.reduction = diffusion_comp,
                                          dims =  1:ncol(diffusion_comp))

qc_boxplot(mc.obj = cd34.supercell, qc.metrics = "separation")
```
### Inner normalized variance (INV)
We can compute the INV for the MetaCell2 and SuperCell metacells as defined in MetaCell paper.

```{r}
cd34.supercell$INV <- mc_INV(cell.membership = cd34.supercell@misc$cell_membership, sc.obj = cd34.singlecells)
cd34.metacell$INV <- mc_INV(cell.membership = cd34.metacell@misc$cell_membership, sc.obj = cd34.singlecells, group.label = "metacell_name")

```


Comparison of MetaCell2 and SuperCell for compactness (lower the better).

```{r}
cd34.metacell$tool <- "MetaCell2"
cd34.supercell$tool <- "SuperCell"
data.metrics <- rbind(cd34.metacell@meta.data[,c("tool","INV","compactness","separation")],
                      cd34.supercell@meta.data[,c("tool","INV","compactness","separation")])

ggplot(data.metrics,aes(y=compactness,x=tool,fill = tool)) + geom_boxplot()
```
Comparison of MetaCell2 and SuperCell for separation (higher the better).

```{r}
ggplot(data.metrics,aes(y=separation,x=tool,fill = tool)) + geom_boxplot()
```

Comparison of MetaCell2 and SuperCell for INV (lower the better).

```{r}
ggplot(data.metrics,aes(y=INV,x=tool,fill = tool)) + geom_boxplot()

```

