---
title: "Metacell analysis on a continuous dataset"
author: "Leonard Herault, Aurelie Gabriel, Mariia Bilous, David Gfeller"
date: "`r Sys.Date()`"
---

In this example we will analyse differentiating hematopoietic stem cells (Human CD34+).
Note that we retrieved the RNA data from a 10X multiome experiment but we won't analyse the ATAC modality.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up the environment


```{r}
library(reticulate)
conda_env <-  conda_list()[reticulate::conda_list()$name == "MetacellAnalysisToolkit","python"]

Sys.setenv(RETICULATE_PYTHON = conda_env)
```


```{r}
#source("SCimplifyAffinityMatrix.R")
library(SuperCell)
library(Seurat)
library(anndata)
library(MetacellAnalysisToolkit)
library(ggplot2)
```
## Metacell building
Here we will use the MCAT bash command line to build the metacells with MetaCell2 and SuperCell. 

### Building metacell with MetaCell2
We use MetaCell2 (`-t MetaCell`) to identify metacells at a gamma of 50 (`-g 50`). We specify that we want a Seurat .rds object in ouput (`-s seurat`).
```{bash}
../MetacellAnalysisToolkit/cli/MCAT -t MetaCell -i ../MetacellAnalysisToolkit/get_data/cd34_multiome_rna.h5ad -o MCAT_output/MetaCell/cd34/ -g 50 -s seurat
```



### Building metacell with SuperCell
We use SuperCell (`-t SuperCell`) to identify metacells at a gamma of 50 (`-g 50`). We specify that we want a Seurat .rds object in ouput (`-s seurat`).
We use 2000 highly variable genes (HVGs,  `-f 2000`) to compute a PCA from which we use 50 components (`-n 50`) to make a k = 30 knn (`-k 30`) graph on which we identify the metacells.
```{bash}
../MetacellAnalysisToolkit/cli/MCAT -t SuperCell -i ../MetacellAnalysisToolkit/get_data/cd34_multiome_rna.h5ad -o MCAT_output/SuperCell/cd34/ -f 2000 -g 50 -n 50 -k 30 -s seurat
```

## Short downstream analysis of the metacells

### MetaCell2 metacells
We load the object obtained with MCAT
```{r}
cd34.metacell <- readRDS("MCAT_output/MetaCell/cd34/mc_Seurat.rds")
cd34.metacell
```
We Normalize the metacells data, identify the HVGs, and we make a classical dimensionality reduction (first a PCA and then vizualisation with a UMAP).
```{r}
cd34.metacell <- NormalizeData(cd34.metacell)
cd34.metacell <- FindVariableFeatures(cd34.metacell)
cd34.metacell <- ScaleData(cd34.metacell)
cd34.metacell <- RunPCA(cd34.metacell)
cd34.metacell <- RunUMAP(cd34.metacell,dims = c(1:50),min.dist = 0.5)
```

Plot the results using Seurat.
```{r}
umap.metacell <- UMAPPlot(cd34.metacell,group.by = "celltype",label = T)
umap.metacell
```
We correctly retrieve the two main differentiation path of CD34+ cells.
In one hand, Megakaryocyte Erythrocyte Progenitor (MEP) fate with GATA2 and then GATA1 transcription factor expression.
On the other, Lymphoid Myeloid Multipotent Progenitor (LMPP) fate with SPI1 transcription factor expression.
```{r}
genes <- c("CD34","GATA2","GATA1","SPI1")
marker.metacell.umaps <- lapply(genes, FUN = function(x){
  plot <- FeaturePlot(cd34.metacell, features = x)
  }
)
marker.metacell.umaps
```
When analyzing metacells it's a good idea to plot their size on these. We can do it with ggplot.
```{r}
data <- cbind(umap.metacell$data,cd34.metacell$size)
colnames(data)[length(colnames(data))] <- "size"
ggplot(data,aes(x= UMAP_1,y=UMAP_2,color = celltype)) + geom_point(aes(size=size)) + theme_classic()
```
```{r}
lapply(marker.metacell.umaps,FUN = function(x){
  data <- cbind(x$data,cd34.metacell$size)
  colnames(data)[length(colnames(data))] <- "size"
  gene <- colnames(data)[length(colnames(data))-1]
  ggplot(data,aes(x= UMAP_1,y=UMAP_2,size=size,colour = .data[[gene]])) + geom_point() + 
    scale_color_gradientn(colors =  c("lightgrey", "blue"), 
                          guide = "colorbar")  +
  theme_classic() + ggtitle(gene)
} )

```
### SuperCell metacells

We can do the same with the metacells obtained with SuperCell

We load the object obtained with MCAT
```{r}
cd34.supercell <- readRDS("MCAT_output/SuperCell/cd34/mc_Seurat.rds")
cd34.supercell
```
We Normalize the supercells data, identify the HVGs, and we make a classical dimensionality reduction (first a PCA and then vizualisation with a UMAP).
```{r}
cd34.supercell <- NormalizeData(cd34.supercell)
cd34.supercell <- FindVariableFeatures(cd34.supercell)
cd34.supercell <- ScaleData(cd34.supercell)
cd34.supercell <- RunPCA(cd34.supercell)
cd34.supercell <- RunUMAP(cd34.supercell,dims = c(1:50),min.dist = 0.5)
```

Plot the results using Seurat.
```{r}
umap.supercell <- UMAPPlot(cd34.supercell,group.by = "celltype",label = T)
umap.supercell
```
We correctly retrieve the two main differentiation path of CD34+ cells.
In one hand, Megakaryocyte Erythrocyte Progenitor (MEP) fate with GATA2 and then GATA1 transcription factor expression.
On the other, Lymphoid Myeloid Multipotent Progenitor (LMPP) fate with SPI1 transcription factor expression.
```{r}
genes <- c("CD34","GATA2","GATA1","SPI1")
marker.supercell.umaps <- lapply(genes, FUN = function(x){
  plot <- FeaturePlot(cd34.supercell, features = x)
  }
)
marker.supercell.umaps
```
When analyzing supercells it's a good idea to plot their size on these. We can do it with ggplot.
```{r}
data <- cbind(umap.supercell$data,cd34.supercell$size)
colnames(data)[length(colnames(data))] <- "size"
ggplot(data,aes(x= UMAP_1,y=UMAP_2,color = celltype)) + geom_point(aes(size=size)) + theme_classic()
```

```{r}
lapply(marker.supercell.umaps,FUN = function(x){
  data <- cbind(x$data,cd34.supercell$size)
  colnames(data)[length(colnames(data))] <- "size"
  gene <- colnames(data)[length(colnames(data))-1]
  ggplot(data,aes(x= UMAP_1,y=UMAP_2,size=size,colour = .data[[gene]])) + geom_point() + 
    scale_color_gradientn(colors =  c("lightgrey", "blue"), 
                          guide = "colorbar")  +
  theme_classic() + ggtitle(gene)
} )

```

## Quality control with the MetacellAnalysisToolkit package

### Loading single-cell data

First we need to load the single cell data and make a classical dimentionality reduction analysis.
```{r}
cd34.singlecells <- read_h5ad("../MetacellAnalysisToolkit/get_data/cd34_multiome_rna.h5ad")
cd34.singlecells <- CreateSeuratObject(counts = Matrix::t(cd34.singlecells$X),meta.data = cd34.singlecells$obs)
cd34.singlecells <- NormalizeData(cd34.singlecells)
cd34.singlecells <- FindVariableFeatures(cd34.singlecells)
cd34.singlecells <- ScaleData(cd34.singlecells)
cd34.singlecells <- RunPCA(cd34.singlecells)
cd34.singlecells <- RunUMAP(cd34.singlecells,dims = c(1:50))
```

```{r}
UMAPPlot(cd34.singlecells,group.by = "celltype",label = T)
```
### Visualization of metacells in single-cell space

Now we can plot metacells from MetaCell2 in the single cell space.
```{r}
mc_projection(
  sc.obj = cd34.singlecells,
  mc.obj = cd34.metacell,
  cell.membership = cd34.metacell@misc$membership,
  sc.reduction = "umap",
  sc.label = "celltype", # single cells will be colored according the sc.label
  metacell.label = "celltype" # metacells cell will be colored according the metacell.label
  ) + theme_classic()
```
And same for metacells obtained with SuperCell.
```{r}
mc_projection(
  sc.obj = cd34.singlecells,
  mc.obj = cd34.supercell,
  cell.membership = cd34.supercell@misc$cell_membership,
  sc.reduction = "umap",
  sc.label = "celltype", # single cells will be colored according the sc.label
  metacell.label = "celltype" # metacells cell will be colored according the metacell.label
  ) + theme_classic()
```
### Compactness and separation

We can compute the compactness and separation of the metacells from the single cell pca. 
With the `mc_compactnness` we can use the approach of SEACells by computing these metrics in a diffusion map obtained from the pca.

First for MetaCell2 metacells




```{r}
cd34.metacell$compactness <- mc_compactness(cell.membership = cd34.metacell@misc$cell_membership, sc.obj = cd34.singlecells,
                                      sc.reduction = "pca", n.components = 50, diffusion.components = T)

## Computing compactness ...

qc_boxplot(mc.obj = cd34.metacell, qc.metrics = "compactness")
```

```{r}
cd34.metacell$separation <- mc_separation(cell.membership = cd34.metacell@misc$cell_membership, sc.obj = cd34.singlecells,
                                      sc.reduction = "pca", n.components = 50, diffusion.components = T)

## Computing compactness ...

qc_boxplot(mc.obj = cd34.metacell, qc.metrics = "separation")
```
Same for metacell obtained with SuperCell.
```{r}
cd34.supercell$compactness <- mc_compactness(cell.membership = cd34.supercell@misc$cell_membership, sc.obj = cd34.singlecells,
                                      sc.reduction = "pca", n.components = 50, diffusion.components = T)

## Computing compactness ...

qc_boxplot(mc.obj = cd34.supercell, qc.metrics = "compactness")
```

```{r}
cd34.supercell$separation <- mc_separation(cell.membership = cd34.supercell@misc$cell_membership, sc.obj = cd34.singlecells,
                                      sc.reduction = "pca", n.components = 50, diffusion.components = T)

## Computing compactness ...

qc_boxplot(mc.obj = cd34.supercell, qc.metrics = "separation")
```
### Inner normalized variance (INV)
We can compute the INV for the metacells as defined in MetaCell paper.

```{r}
cd34.supercell$INV <- mc_INV(cell.membership = cd34.supercell@misc$cell_membership, sc.obj = cd34.singlecells)
cd34.metacell$INV <- mc_INV(cell.membership = cd34.metacell@misc$cell_membership, sc.obj = cd34.singlecells)

```


Comparison of MetaCell2 and SuperCell for compactness (lower the better).

```{r}
cd34.metacell$tool <- "MetaCell2"
cd34.supercell$tool <- "SuperCell"
data.metrics <- rbind(cd34.metacell@meta.data[,c("tool","INV","compactness","separation")],cd34.supercell@meta.data[,c("tool","INV","compactness","separation")])

ggplot(data.metrics,aes(y=compactness,x=tool,fill = tool)) + geom_boxplot()
```
Comparison of MetaCell2 and SuperCell for separation (higher the better).

```{r}
ggplot(data.metrics,aes(y=separation,x=tool,fill = tool)) + geom_boxplot()
```

Comparison of MetaCell2 and SuperCell for INV (lower the better).

```{r}
ggplot(data.metrics,aes(y=INV,x=tool,fill = tool)) + geom_boxplot()

```

