--- 
title: "Metacell Tutorial"
author: "Aurélie Gabriel, Léonard Hérault, Mariia Bilous, David Gfeller"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [citations.bib, packages.bib]
description: |
  This is a tutorial about metacell construction and analysis.
link-citations: yes
github-repo: GfellerLab/Metacell_tutorial
biblio-style: apalike
csl: chicago-fullnote-bibliography.csl
---

# This tutorial{.unnumbered}

In this tutorial, we describe the different steps that should be followed to build metacells[@baran_metacell_2019] from single-cell data using three frameworks:
SuperCell[@SuperCell] (tutorial in \@ref(SuperCell-construction)), (ii) Metacells version 2 (MC2[@MC2]) (tutorial in \@ref(MC2-construction)), and (iii) SEACells[@SEACells] (tutorial in \@ref(SEACells-construction)).

We also show how to obtain metacells by running these methods using a command line tool that we provide as part of the MetacellAnalysisToolkit (MCAT) [github repository](https://github.com/GfellerLab/MetacellToolkit).
This repository also contains the MetacellAnalysisToolkit R package which provides R functions to compute QC
metrics and visualization functions to evaluate the quality of metacells.
In chapter \@ref(XXX), we use MCAT to build metacells on a continuous dataset of CD34+ cells and use the R package to visualize the constructed metacells.
In chapter \@ref(QCs), we describe how to use the R package to evaluate the quality of the metacells.

Finally, we provide examples of downstream analyses performed at the metacell level. These analyses include clustering, differential analysis, data integration and gene regulatory network analysis.

```{r include=FALSE}
source('./R/config.R')
```

```{r include=FALSE}
TO_CACHE = FALSE
```
# Requirements 

This chapter describes how to obtain the packages and data needed to reproduce the analyses performed in this tutorial.

## Installations {#installations}

### Using conda (recommended)
To build a conda environment containing the three metacell building tools used in this tutorial (SuperCell, MC2 and SEACells), 
please follow the instructions provided in the README of our MetacellAnalysisToolkit [github repository](https://github.com/GfellerLab/MetacellToolkit).


```{r, eval = TRUE}
library(reticulate)
conda_env <-  conda_list()[reticulate::conda_list()$name == "MetacellAnalysisToolkit","python"]

use_condaenv(conda_env)
```

### Without conda
If you don't have conda, you can use the following instructions:

Set up a python virtual environment with MC2 and SEACells installed:

```{bash, eval = FALSE}
pip install virtualenv
virtualenv my_env
source my_env/bin/activate

# Installing SEACells
git clone https://github.com/dpeerlab/SEACells.git
cd SEACells
python setup.py install
cd ..
pip install -r SEACells_requirements.txt
pip install ipywidgets
pip install jupyter

# Install MC2
pip install git+https://github.com/tanaylab/metacells
```

In R, install the SuperCell package:
```{r, eval = FALSE, echo = TRUE}
remotes::install_github("GfellerLab/SuperCell", force = TRUE, upgrade = FALSE)
```

To run python function in R, install reticulate:
```{r, eval = FALSE, echo = TRUE}
install.packages('reticulate')
```

To use the python libraries installed in the virtual environment, define the RETICULATE_PYTHON variable as follow:
```{bash, eval = FALSE, echo = TRUE}
echo 'RETICULATE_PYTHON=my_env/bin/python' > '.Renviron'
```

## Retrieve a discrete dataset (PBMCs dataset) {#PBMC-data}

To test metacell construction on a discrete dataset, we retrieved the 3k PBMCs from scanpy datasets as follows:
```{python, eval = T, collapse = T, cache = TO_CACHE}
import scanpy as sc 
import os

adata = sc.datasets.pbmc3k()
adata_proc = sc.datasets.pbmc3k_processed()

adata       = adata[adata_proc.obs_names].copy()
adata.obs   = adata_proc.obs.copy()
adata.uns   = adata_proc.uns.copy()
adata.obsm  = adata_proc.obsm.copy()
adata.obsp  = adata_proc.obsp.copy()

adata.X = adata.X.astype("float32")
raw_ad = sc.AnnData(adata.X.copy())
raw_ad.obs_names, raw_ad.var_names = adata.obs_names, adata.var_names
adata.raw = raw_ad

```

The data are saved in the following file for future analyses in python (use of SEACells and MC2): "data/3k_pbmc/singlecell_anndata_filtered.h5ad".

```{python, eval = T, collapse = T, cache = TO_CACHE}
directory = os.path.join("data", "3k_pbmc")

if not os.path.exists(directory):
    os.makedirs(directory)
  
adata.write_h5ad(os.path.join("data", "3k_pbmc", "singlecell_anndata_filtered.h5ad"))
```

The data are saved in the following file for future analyses in R (use of SuperCell): "data/3k_pbmc/singlecell_seurat_filtered.rds".

```{r, eval = T, collapse = T, cache = TO_CACHE}
library(reticulate)
library(Seurat)
library(anndata)
adata <- anndata::read_h5ad(file.path("data/3k_pbmc/singlecell_anndata_filtered.h5ad"))

raw_counts <- Matrix::t(adata$raw$X)
colnames(raw_counts) <- rownames(adata$obs)
rownames(raw_counts) <- rownames(adata$var)

pbmc <- CreateSeuratObject(counts = raw_counts, meta.data = adata$obs)
saveRDS(pbmc, file = paste0("data/3k_pbmc/singlecell_seurat_filtered.rds"))
```


## Retrieve a continuous dataset (CD34 dataset) {#CD34-data}

To test metacell construction on discrete dataset, we retrieved the CD34 dataset provided by Persad et al. [@SEACells]:
```{bash, eval = FALSE, cache = TO_CACHE}
mkdir data/CD34
wget -O data/CD34/cd34_multiome_rna.h5ad 'https://zenodo.org/record/6383269/files/cd34_multiome_rna.h5ad?download=1' 
```

```{python, eval = T, collapse = T, cache = TO_CACHE}
import scanpy as sc 
import os

adata = sc.read(os.path.join("data", "CD34", "cd34_multiome_rna.h5ad"))
adata.X.sort_indices()
raw_ad = sc.AnnData(adata.X.copy())
raw_ad.obs_names, raw_ad.var_names = adata.obs_names, adata.var_names
adata.raw = raw_ad

sc.pl.embedding(adata, 'X_umap', color='celltype')
```

The data are saved in the following file for future analyses in python (use of SEACells and MC2): "data/CD34/singlecell_anndata_filtered.h5ad".
```{python, eval = T, collapse = T, cache = TO_CACHE}
directory = os.path.join("data", "cd34_multiome")

if not os.path.exists(directory):
    os.makedirs(directory)
    
adata.write_h5ad(os.path.join("data", "CD34", "singlecell_anndata_filtered.h5ad"))
```

The data are saved in the following file for future analyses in R (use of SuperCell): "data/CD34/singlecell_seurat_filtered.rds".

```{r, eval = T, collapse = T, cache = TO_CACHE}
library(reticulate)
library(Seurat)
library(anndata)
adata <- anndata::read_h5ad(file.path("data/CD34/singlecell_anndata_filtered.h5ad"))

raw_counts <- Matrix::t(as(adata$raw$X, "CsparseMatrix"))
colnames(raw_counts) <- rownames(adata$obs)
rownames(raw_counts) <- rownames(adata$var)

cd34 <- CreateSeuratObject(counts = raw_counts, meta.data = adata$obs)
saveRDS(cd34, file = file.path("data/CD34/singlecell_seurat_filtered.rds"))
```

## Retrieve the lung atlas dataset {#HLCA-data}

### Downloading the atlas

To illustrate how metacells can be used in the context of single-cell data integration,
we used a cell atlas of the human lung (core) available on [cellxgene](https://cellxgene.cziscience.com/collections/6f6d381a-7701-4781-935c-db10d30de293). 
To download the data, please choose the `.h5ad` option after clicking on the download button for the core atlas (3 tissues, 584'944 cells).

Save these data in the `data/HLCA/` directory. 

Please note that this may take some time (\~45 mins) as the file is quite large (5.6 GB).

###  Splitting atlas by datasets

We will use anndata to read in backed mode (saving a lot of memory) the whole atlas and write one h5ad file for each dataset. 
This should take less than 10 minutes.

If you are limited in time feel free to process only a subset of the dataset.

```{r , eval = FALSE, collapse = T, cache = TO_CACHE}
t0.split <- Sys.time()

library(anndata)
adata <- read_h5ad("data/HLCA/local.h5ad",backed = "r")
adata$var_names <- adata$var$feature_name # We will use gene short name for downstream analyses
datasets <- unique(adata$obs$dat)

# If you are limited in time you can process on half of the datasets (uncomment th following line)
# datasets <- datasets[1:7]

print(dim(adata))

lapply(datasets,FUN =  function(x) {
  dir.create(paste0("data/HLCA/datasets/",x),recursive = T)
  adata.dataset <- AnnData(X = adata[adata$obs$dataset == x]$raw$X,
                           var = adata[adata$obs$dataset == x]$var,
                           obs = adata[adata$obs$dataset == x]$obs)
  #This will allow us to construct supervised metacell for each cell type in each sample later in the tutorial
  adata.dataset$obs$ann <- as.character(adata.dataset$obs$ann_level_3)
  # For cell without an annotation at the 3rd level we will use the second level of annotation
  adata.dataset$obs$ann[adata.dataset$obs$ann_level_3 == 'None'] = as.character(adata.dataset$obs$ann_level_2[adata.dataset$obs$ann_level_3 == 'None'])
  adata.dataset$obs$ann_sample <- paste0(adata.dataset$obs$ann,"_",adata.dataset$obs$sample)
  
  write_h5ad(adata.dataset,paste0("data/HLCA/datasets/",x,"/sc_adata.h5ad"))
}
)

remove(adata)
gc()

tf.split <- Sys.time()
tf.split - t0.split
```
