
[//]: # (Chunk to run MC-2 metacell construction for a discrete -pbmcs- dataset)

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

In this section, we construct metacells using [Metacell-2 (MC2)](https://github.com/tanaylab/metacells) implemented in Python. 

### Method 
Metacell-2 (MC2) is a python tool to construct metacells and is the updated version of the MetaCell algorithm, which introduced the concept of metacell.
MC2 applies a two-phase divide-and-conquer approach. Cells are randomly divided into piles of ~10k cells and 
initial metacells are built applying a MetaCell-like approach per pile, i.e. based on a single-cell kNN graph built 
from log-normalized counts using a set of highly variable genes. 
Then, transcriptionally similar metacells are grouped into metagroup piles for the identification of final metacells and outliers identification.
Note that prior to metacell identification, the MC2 framework recommends gene filtering steps. The choice of the genes used
by the method is of high importance to assure good quality of the metacells.

The code provided in this section is adapted from the [author's tutorial](https://tanaylab.github.io/metacells-vignettes/one-pass.html). 
For more information on the method, please refer to our review @Review and the original paper [@MC2]. 

#### Importing python packages {-}

To run MC2, the following python packages need to be imported: 

```{python import-mc2, eval = FALSE}
import os
import numpy as np
import pandas as pd
import anndata as ad
import scanpy as sc
import matplotlib.pyplot as plt
import seaborn as sns
import metacells as mc
```


If you don't have these packages installed, please refer to the section \@ref(installations).

```{python mc2-parameters, eval = TRUE, echo = FALSE}
MC_tool = "MC2"

## Here we can modify dataset
proj_name = ["cell_lines", "3k_pbmc", "bmcite"][2]

annotation_label = {'cell_lines':'cell_line',
                   '3k_pbmc':'louvain',
                   'bmcite':'celltype_simplified'}[proj_name] # name of annotation field (obs)
```

### Data loading 
We will run Metacell-2 (MC2) on a single-cell dataset composed of around 30'000 bone marrow cells. 
Please follow the section \@ref(bmcite-data) to retrieve these data from the SeuratData package and save the data in the following file: "data/bmcite/singlecell_anndata_filtered.h5ad".
```{r load-data, child='./functional_chunks/load_anndata.Rmd'}
```

```{python load-data-noEval, eval = FALSE, echo = TRUE}
MC_tool = "MC2"
proj_name = "bmcite"
annotation_label = 'celltype_simplified'
ad = sc.read(os.path.join("data", proj_name, "singlecell_anndata_filtered.h5ad"))
ad.var.index = ad.var.genes
```

We initialize the name of the anndata (in the unstructured annotations) object using the `mc.ut.set_name()` function from the MC2 package.

```{python}
mc.ut.set_name(ad, proj_name)
```

### Filtering steps 
MC2 requires that standard filtering steps such as doublet filtering are performed outside of the MC2 framework. 
In addition to standard data filtering steps, the MC2 package proposes functions to filter the single-cell data at the gene and at the cell level (See [author's vignette](https://tanaylab.github.io/metacells-vignettes/one-pass.html)). 
At the gene level, the filtering steps consist in excluding genes based on biological knowledge (*e.g.* mitochrondrial genes) as well as based on their expression levels. 
The latter genes include genes with zero expression or low expression levels and "bursty lonely genes" (*i.e.*, genes with high expression levels but no correlation with any other gene). 
At the cell level, filtering is performed based on cells UMI counts.

#### Gene filtering {-}

In the following code chunk, we exclude genes using the `mc.pl.exclude_genes()`function from the MC2 package. 
Based on the authors vignette, we consider a minimal list of genes to exclude, *i.e.*, sex-specific and non-coding genes as well as the mitochondrial genes. 
To complete this list of genes, an iterative approach can be used following the guidelines of the MC2 authors in a [second vignette](https://tanaylab.github.io/metacells-vignettes/iterative.html).
The `mc.pl.exclude_genes()` function will filter out: 
i) the known-to-be-excluded genes defined by the user as gene names or gene names patterns (`EXCLUDED_GENE_NAMES` and `EXCLUDED_GENE_PATTERNS` parameters respectively), 
and ii) the "bursty lonely genes". 

```{python mc2-gene-filtering, collapse = TRUE, result = 'hold'}
EXCLUDED_GENE_NAMES = ["XIST", "MALAT1", "NEAT1"] 
EXCLUDED_GENE_PATTERNS = ['MT-.*']

mc.pl.exclude_genes(
    ad,
    excluded_gene_names=EXCLUDED_GENE_NAMES,
    excluded_gene_patterns=EXCLUDED_GENE_PATTERNS,
    random_seed=123456
)

```

#### Cell filtering based on UMIs counts {-}

In the MC2 framework, cells with very low and very high UMI content will be filtered out (`PROPERLY_SAMPLED_MIN_CELL_TOTAL`, `PROPERLY_SAMPLED_MAX_CELL_TOTAL` variables defining thresholds in the next code chunk).  
Also, cell filtering based on UMI counts in excluded genes is also performed(`PROPERLY_SAMPLED_MAX_EXCLUDED_GENES_FRACTION` variable). 
Since our dataset has been pre-filtered, very lenient cutoffs will be used in this tutorial. 
The following code chunk defines these parameters. 
To adapt them to your datasets, we advise you to explore the distributions of total UMI counts and UMI counts in excluded genes, as recommended and described in the MC2 [original vignette](https://tanaylab.github.io/metacells-vignettes/one-pass.html).

```{python mc2-cell-filtering-parameters, warning = FALSE, collapse = TRUE, result = 'hold'}
PROPERLY_SAMPLED_MIN_CELL_TOTAL = 200 
PROPERLY_SAMPLED_MAX_CELL_TOTAL = 10000 
PROPERLY_SAMPLED_MAX_EXCLUDED_GENES_FRACTION = 0.25
```

The number of UMIs in excluded genes is computed using the `mc.tl.compute_excluded_gene_umis()` function and cells are filtered out using the `mc.pl.exclude_cells()` function.
Additional cells can be filtered out by adding a cell description columns in the `obs` data frame in the anndata oject. This annotation should be a boolean indicating whether the cell should filtered out or not. 
The name of this column should be provided to the `mc.pl.exclude_cells()` function via the `additional_cells_masks` parameter. 

```{python mc2-cells-filtering, result='hold', collapse = TRUE}
mc.tl.compute_excluded_gene_umis(ad)

mc.pl.exclude_cells(
    ad,
    properly_sampled_min_cell_total=PROPERLY_SAMPLED_MIN_CELL_TOTAL,
    properly_sampled_max_cell_total=PROPERLY_SAMPLED_MAX_CELL_TOTAL,
    properly_sampled_max_excluded_genes_fraction=PROPERLY_SAMPLED_MAX_EXCLUDED_GENES_FRACTION # ,
    # additional_cells_masks=["|doublet_cell"]
)
```

After performing the two-step filtering (genes and cells), the "cleaned" data can be extracted using the `mc.pl.extract_clean_data()` function.

```{python mc-extract-clean-cells, result='hold', collapse = TRUE}
# Extract clean dataset (with filtered cells and genes)
ad = mc.pl.extract_clean_data(ad)
```


### Building metacells

#### Defining lateral genes {-}

To build metacells, we need to define lateral genes, which are genes with strong biological signal which is independent of cell-state, *e.g.* cell-cycle genes. 
These genes will be ignored for computing cells similarity and building metacells 
but will be considered to define outlier cells (*i.e.*, expression levels of lateral genes should be consistent within metacells).
In the following chunk, we consider a minimal list of lateral genes (provided by the MC2 authors) including cell-cycle and ribosomal genes and 
mark them in the MC2 object using the `mc.pl.mark_lateral_genes()` function.

```{python mc-lateral-genes, result='hold', collapse = TRUE}

LATERAL_GENE_NAMES = [
    "ACSM3", "ANP32B", "APOE", "AURKA", "B2M", "BIRC5", "BTG2", "CALM1", "CD63", "CD69", "CDK4",
    "CENPF", "CENPU", "CENPW", "CH17-373J23.1", "CKS1B", "CKS2", "COX4I1", "CXCR4", "DNAJB1",
    "DONSON", "DUSP1", "DUT", "EEF1A1", "EEF1B2", "EIF3E", "EMP3", "FKBP4", "FOS", "FOSB", "FTH1",
    "G0S2", "GGH", "GLTSCR2", "GMNN", "GNB2L1", "GPR183", "H2AFZ", "H3F3B", "HBM", "HIST1H1C",
    "HIST1H2AC", "HIST1H2BG", "HIST1H4C", "HLA-A", "HLA-B", "HLA-C", "HLA-DMA", "HLA-DMB",
    "HLA-DPA1", "HLA-DPB1", "HLA-DQA1", "HLA-DQB1", "HLA-DRA", "HLA-DRB1", "HLA-E", "HLA-F", "HMGA1",
    "HMGB1", "HMGB2", "HMGB3", "HMGN2", "HNRNPAB", "HSP90AA1", "HSP90AB1", "HSPA1A", "HSPA1B",
    "HSPA6", "HSPD1", "HSPE1", "HSPH1", "ID2", "IER2", "IGHA1", "IGHA2", "IGHD", "IGHG1", "IGHG2",
    "IGHG3", "IGHG4", "IGHM", "IGKC", "IGKV1-12", "IGKV1-39", "IGKV1-5", "IGKV3-15", "IGKV4-1",
    "IGLC2", "IGLC3", "IGLC6", "IGLC7", "IGLL1", "IGLL5", "IGLV2-34", "JUN", "JUNB", "KIAA0101",
    "LEPROTL1", "LGALS1", "LINC01206", "LTB", "MCM3", "MCM4", "MCM7", "MKI67", "MT2A", "MYL12A",
    "MYL6", "NASP", "NFKBIA", "NUSAP1", "PA2G4", "PCNA", "PDLIM1", "PLK3", "PPP1R15A", "PTMA",
    "PTTG1", "RAN", "RANBP1", "RGCC", "RGS1", "RGS2", "RGS3", "RP11-1143G9.4", "RP11-160E2.6",
    "RP11-53B5.1", "RP11-620J15.3", "RP5-1025A1.3", "RP5-1171I10.5", "RPS10", "RPS10-NUDT3", "RPS11",
    "RPS12", "RPS13", "RPS14", "RPS15", "RPS15A", "RPS16", "RPS17", "RPS18", "RPS19", "RPS19BP1",
    "RPS2", "RPS20", "RPS21", "RPS23", "RPS24", "RPS25", "RPS26", "RPS27", "RPS27A", "RPS27L",
    "RPS28", "RPS29", "RPS3", "RPS3A", "RPS4X", "RPS4Y1", "RPS4Y2", "RPS5", "RPS6", "RPS6KA1",
    "RPS6KA2", "RPS6KA2-AS1", "RPS6KA3", "RPS6KA4", "RPS6KA5", "RPS6KA6", "RPS6KB1", "RPS6KB2",
    "RPS6KC1", "RPS6KL1", "RPS7", "RPS8", "RPS9", "RPSA", "RRM2", "SMC4", "SRGN", "SRSF7", "STMN1",
    "TK1", "TMSB4X", "TOP2A", "TPX2", "TSC22D3", "TUBA1A", "TUBA1B", "TUBB", "TUBB4B", "TXN", "TYMS",
    "UBA52", "UBC", "UBE2C", "UHRF1", "YBX1", "YPEL5", "ZFP36", "ZWINT"
]
LATERAL_GENE_PATTERNS = ["RP[LS].*"]  # Ribosomal

# This will mark as "lateral_gene" any genes that match the above, if they exist in the clean dataset.
mc.pl.mark_lateral_genes(
    ad,
    lateral_gene_names=LATERAL_GENE_NAMES,
    lateral_gene_patterns=LATERAL_GENE_PATTERNS,
)

```

Some genes have higher variances than expected which could lead to false positive outlier identification.
Users can mark those genes as *noisy genes* using the `mc.pl.mark_noisy_genes()` function. 
```{python mc-noisy-genes, result='hold', collapse = TRUE}

NOISY_GENE_NAMES = [
    "CCL3", "CCL4", "CCL5", "CXCL8", "DUSP1", "FOS", "G0S2", "HBB", "HIST1H4C", "IER2", "IGKC",
    "IGLC2", "JUN", "JUNB", "KLRB1", "MT2A", "RPS26", "RPS4Y1", "TRBC1", "TUBA1B", "TUBB"
]
# This will mark as "noisy_gene" any genes that match the above, if they exist in the clean dataset.
mc.pl.mark_noisy_genes(ad, noisy_gene_names=NOISY_GENE_NAMES)

```

To extend this list of lateral genes, users can use the `relate_to_lateral_genes` function to identify genes that are highly correlated with the predefined lateral genes.
The use of this function is described in [the vignette from the MC2 authors](https://tanaylab.github.io/metacells-vignettes/iterative.html).

#### Define target_metacell_size (graining level) {-}

By default, MC2 will build metacells with a size of 96 cells per metacells. 
Users can vary the `target_metacell_size` parameter to reach a desired graining level. 

```{python estimate-target-mc-size}
gamma = 75
target_metacell_size = gamma
```

#### Metacells identification using the divide and conquer approach {-}

The construction of metacells by MC2 is performed using the `mc.pl.divide_and_conquer_pipeline()` function.
Note that by default all cores of the system will be used for the metacells construction. 
To change this behavior and adapt the number of cores the MC2 authors propose to use the `mc.pl.guess_max_parallel_piles()` and `mc.pl.set_max_parallel_piles()` functions 
to adapt the number of processed in parallel depending on the available memory.

The `mc.pl.divide_and_conquer_pipeline()` function associates each cell to a metacell or defines the cell as outlier. 
These assignments are found in the `obs` layer of the anndata object.

```{python mc2-divide-n-conquer, message=FALSE, warning=FALSE, result=FALSE}
max_parallel_piles = mc.pl.guess_max_parallel_piles(ad)
mc.pl.set_max_parallel_piles(max_parallel_piles)
mc.pl.divide_and_conquer_pipeline(
    ad,
    target_metacell_size = target_metacell_size,
    random_seed = 123456)

ad.obs.metacell.head
```


The following code chunk adds a columns (named `membership`) containing the single_cell assignments to the obs attribute in the single-cell anndata object. 
The membership information is required to compute metacells quality metrics as shown in chapter \@ref(QCs). 

```{python mc2-membership, result = FALSE}
# make a membership -- index of metacells to which single cells belong to 
ad.obs['membership'] = [int(i)+1 if i >= 0 else np.nan for i in ad.obs.metacell] 
```

#### Retrieve aggregated metacell data {-}

The function `mc.pl.collect_metacells()` should be used to subsequently retrieve an anndata object containing the data at the metacells level instead of the single-cell level. 
This function will store in the `X` data matrix of the anndata object a matrix of gene fraction (*i.e.*, the sum of all gene levels in a metacell sums to 1)
and it will store the total UMIs per gene per metacell in the layer `total_umis`. 

```{python mc2-aggregation, echo=TRUE}

mc_ad = mc.pl.collect_metacells(ad, name='metacells', random_seed = 123456)
mc_ad.shape
mc_ad.X.sum(axis=1)[:5] 
mc_ad.layers['total_umis']

```

### Annotate metacells (using available annotations)

If single-cell annotations are available in the original single-cell anndata object. We can transfer these annotations to the metacell anndata object
using the `mc.tl.convey_obs_to_group()` function which will associate each metacell to the most frequent annotation (categorical) or 
averaged annotation (continuous) across the single-cells composing the metacell 
(use of the `mc.ut.most_frequent` and `np.mean` respectively in the `mode` paratemer). 
 
```{python mc2-annotation, result = FALSE}
# Assign a single value for each metacell based on the cells.
mc.tl.convey_obs_to_group(
    adata=ad, gdata=mc_ad,
    property_name=annotation_label, to_property_name=annotation_label,
    method=mc.ut.most_frequent  # This is the default, for categorical data
)

# Compute the fraction of cells with each possible value in each metacell:
mc.tl.convey_obs_fractions_to_group(  
    adata=ad, gdata=mc_ad,
    property_name=annotation_label, to_property_name=annotation_label
)

```
 
### Save output 

For future downstream analyses in python (section \@ref(standard-analysis-Py)), we save the metacell counts in an Anndata object: 

```{python mc2-save-py}
## Save single-cell metadata (i.e., `raw.obs` dataframe) in the metacell adata object
mc_ad.uns = ad.uns.copy()
mc_ad.uns['sc.obs'] = ad.obs.copy()

# save the requested gamma
mc_ad.uns['gamma'] = gamma

# save metacell size
mc_ad.obs.rename(columns={'grouped':'size'}, inplace = True)

print("Saving metacell object for the "+ proj_name+ " dataset using "+ MC_tool)
mc_ad.write_h5ad(os.path.join('./data', proj_name, f'metacell_{MC_tool}.h5ad'))
```

For future QCs and downstream analyses in R (section \@ref(standard-analysis-R)), we save the metacell counts in a Seurat object: 
```{r mc2-save-seurat}
library(Seurat)
library(anndata)
library(reticulate)
adata_mc <- read_h5ad(paste0("data/", py$proj_name, "/metacell_MC2.h5ad"))

# Save counts and metadata in a Seurat object
countMatrix <-  Matrix::t(adata_mc$X)
colnames(countMatrix) <- adata_mc$obs_names
rownames(countMatrix) <- adata_mc$var_names
MC.seurat <- CreateSeuratObject(counts = as(countMatrix, 'CsparseMatrix'), meta.data = as.data.frame(adata_mc$obs))
MC.seurat@misc[["var_features"]] <- rownames(adata_mc$var)[which(adata_mc$var$selected_gene == T)] 

# Save membership in misc
MC.seurat@misc$cell_membership <- py$ad$obs['membership']
saveRDS(MC.seurat, file = paste0('./data/', py$proj_name, '/metacell_MC2.rds'))
```

```{python mc2-remove-variables, include = FALSE}
globals().pop('ad', None);
globals().pop('raw', None);
globals().pop('mc_ad', None);
globals().pop('membership', None);
globals().pop('mc_size', None);
globals().pop('mc_purity', None);
```



