
[//]: # (Weighted downstream analysis of metacells using R (Seurat))

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r r-libraries-weighted, eval=TRUE}
library(Seurat) 
library(dplyr)
library(ggplot2)
library(SuperCell)
```

### Load metacell Seurat object

We will use Seurat objects containing the metacells counts data and their annotation (*e.g.* and cell-type annotation) and
proceed with downstream analyses considering the size of each metacells.
Seurat objects containing metacells counts data and their annotation were generated at the end of sections \@ref(SuperCell-construction)
These objects can also be generated using the command line described in chapter \@ref(command-line)

```{r r-mc-loading-weighted, eval=TRUE}
MC_tool = "SuperCell"
proj_name = "3k_pbmc"
annotation_column = "louvain"
celltype_colors <- c(
  "CD14+ Monocytes"    = "#E69F00",  # orange
  "B cells"            = "#56B4E9",  # sky blue
  "CD4 T cells"        = "#009E73",  # bluish green
  "NK cells"           = "#F0E442",  # yellow
  "CD8 T cells"        = "#0072B2",  # blue
  "FCGR3A+ Monocytes"  = "#D55E00",  # vermillion
  "Dendritic cells"    = "#CC79A7",  # reddish purple
  "Megakaryocytes"     = "#000000"   # black
)
MC.seurat = readRDS(paste0('./data/', proj_name, '/metacell_', MC_tool,'.rds'))
```


### Dimensionality reduction

As for single-cells, we normalize the raw counts (here aggregated raw counts) and we identify the most variable features in the metacells gene expression data.
Based on these features, we run a sample weighted PCA using the function `supercell_prcomp` from the SuperCell R package
and use the first principal components to obtain a two dimensionnal representation of the data using UMAP.
Using the `supercell_DimPlot` function from the the SuperCell R package we can visualize the metacells and their sized in UMAP space.

```{r r-mc-dim-reduc-weighted, eval=TRUE}
MC.seurat <- NormalizeData(MC.seurat, normalization.method = "LogNormalize")

MC_list <- list(N.SC = ncol(MC.seurat),
                supercell_size = MC.seurat$size)
MC_list$PCA <- SuperCell::supercell_prcomp(
  Matrix::t(GetAssayData(MC.seurat, slot = "data")),
  genes.use = MC.seurat@misc$var_features,  # or a new set of HVG can be computed
  supercell_size = MC_list$supercell_size, # provide this parameter to run sample-weighted version of PCA,
  k = 30
)

MC_list$UMAP <- supercell_UMAP(
  SC = MC_list,
  PCA_name = "PCA",
  n.comp = 30, n_neighbors = 15
)

supercell_DimPlot(SC = MC_list, 
  groups = MC.seurat@meta.data[, annotation_column],
  dim.name = "UMAP", 
  title = paste0("UMAP of metacells colored by cell type assignment"), color.use = celltype_colors
)

```

### Clustering

We cluster the metacells using the function `supercell_cluster` from SuperCell R package to perform the clustering step and visualize these clusters in the UMAP space:
```{r r-mc-clustering-weighted, eval=TRUE}
# compute distance among metacells
D  <- dist(MC_list$PCA$x)

# cluster metacells
MC_list$SCclustering  <- supercell_cluster(D = D, k = 8, supercell_size = MC_list$supercell_size)
MC.seurat$SCclustering <- MC_list$SCclustering$clustering

# Plot clustering result
supercell_DimPlot(
  MC_list, 
  groups = factor(MC_list$SCclustering$clustering),
  dim.name = "UMAP", 
  title = paste0("UMAP of metacells colored by metacell clustering")
)
```

### Differential expression analysis

We perform diffrential analysis to identify the markers of our clusters using the `supercell_FindAllMarkers` function from the SuperCell package. 
```{r r-mc-DEG-weighted, eval=TRUE}
# Compute upregulated genes in each cell line (versus other cells)
MC.all.markers <- supercell_FindAllMarkers(
  ge = GetAssayData(MC.seurat, slot = "data"), 
  clusters = MC_list$SCclustering$clustering, 
  supercell_size = MC_list$supercell_size,
  only.pos = TRUE, 
  min.pct = 0, 
  logfc.threshold = 0.2
)

```

We select the markers for cluster 4: 
```{r r-mc-topMarkers-weighted, eval=TRUE}
cluster4_markers <- MC.all.markers[[4]]
MC.top.markers <- cluster4_markers[order(cluster4_markers$logFC, decreasing = T),]
```

We visualize the top 5 markers for the cluster 4 and see that the top marker genes for this cluster contain marker genes of natural killer cells such as GZMB and GNLY.
```{r r-mc-plot-genes-weighted, eval=TRUE, fig.width = 10, fig.height = 5}
Idents(MC.seurat) <- "SCclustering"
# genes.to.plot <- MC.seurat.top.markers$gene[MC.seurat.top.markers$cluster == unique(MC.seurat@meta.data[,annotation_column])[1]]
# genes.to.plot <- MC.top.markers$gene[c(seq(1, 20, 5))]
genes.to.plot <- rownames(MC.top.markers)[1:5]
VlnPlot(MC.seurat, features = genes.to.plot, ncol = 5, pt.size = 0.0)  
```


