
[//]: # (Standard downstream analysis of metacells using R (Seurat))

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

In this tutorial, standard analyses include dimensionality reduction, clustering and differential expression using the [Seurat](https://satijalab.org/seurat/) framework.

```{r r-libraries, eval=TRUE}
library(Seurat) 
library(dplyr)
library(ggplot2)
```

### Load metacell Seurat object

We will use Seurat objects containing the metacells counts data and their annotation (*e.g.* cell-type annotation) and
proceed with standard Seurat downstream analyses.
Seurat objects containing metacells counts data and their annotation were generated at the end of sections \@ref(SuperCell-construction)
These objects can also be generated using the command line described in chapter \@ref(command-line)

```{r r-mc-loading, eval=TRUE}
MC_tool = "SuperCell"
proj_name = "bmcite"
annotation_column = "celltype_simplified"

cell_types <- c("Prog_RBC", "Unconventional T", "Naive CD4 cell", "Non-Naive CD4 cell", 
                "CD14 Mono", "B cell", "Naive CD8 cell", "Non-Naive CD8 cell", 
                "NK", "GMP", "CD16 Mono", "pDC", "cDC2", "Prog_B 2", 
                "Prog_Mk", "Plasmablast", "HSC", "LMPP", "Prog_DC", "Prog_B 1")

celltype_colors <- c("#7E57C2", "#1E88E5", "#FFC107", "#004D40", "#9E9D24", 
                 "#F06292", "#546E7A", "#D4E157", "#76FF03", "#6D4C41", 
                 "#26A69A", "#AB47BC", "#EC407A", "#D81B60", "#42A5F5", 
                 "#2E7D32", "#FFA726", "#5E35B1", "#EF5350", "#3949AB")
names(celltype_colors) <-cell_types
MC.seurat = readRDS(paste0('./data/', proj_name, '/metacell_', MC_tool,'.rds'))
```


### Dimensionality reduction

As for single-cells, we normalize the raw counts (here aggregated raw counts) and we identify the most variable features in the metacells gene expression data.
Based on these features, we run PCA and use the first principal components to obtain a two dimensionnal representation of the data using UMAP.

```{r r-mc-dim-reduc, eval=TRUE}
Idents(MC.seurat) <- annotation_column
MC.seurat <- NormalizeData(MC.seurat)
MC.seurat <- FindVariableFeatures(MC.seurat, selection.method = "vst", nfeatures = 2000)
MC.seurat <- ScaleData(MC.seurat)
MC.seurat <- RunPCA(MC.seurat, verbose = F)
MC.seurat <- RunUMAP(MC.seurat, dims = 1:30, verbose = F, min.dist = 1)

data <- cbind(Embeddings(MC.seurat, reduction = "umap"),
              data.frame(size = MC.seurat$size,
                         cell_type = MC.seurat@meta.data[, annotation_column]))

p_annot <- ggplot(data, aes(x= UMAP_1, y=UMAP_2, color = cell_type)) + geom_point(aes(size=size)) +
  ggplot2::scale_size_continuous(range = c(0.5,  0.5*max(log((data$size))))) +
   ggplot2::scale_color_manual(values = celltype_colors) +
  theme_classic() + guides(color=guide_legend(ncol=2))
p_annot
#DimPlot(MC.seurat, reduction = "umap", cols = celltype_colors, pt.size = log1p(MC.seurat$size)) 
```

### Clustering

We cluster the metacells using Seurat clustering steps and visualize these clusters using UMAP:
```{r r-mc-clustering, eval=TRUE}
MC.seurat$SCclustering  <- SuperCell::supercell_cluster(D = dist(MC.seurat@reductions$pca@cell.embeddings[, 1:30]  ), k = 15)$clustering
data <- cbind(Embeddings(MC.seurat, reduction = "umap"),
              data.frame(size = MC.seurat$size,
                         cluster = paste0("C_",MC.seurat$SCclustering)))

p_cluster <- ggplot(data, aes(x= UMAP_1, y=UMAP_2, color = cluster)) + geom_point(aes(size=size)) +
  ggplot2::scale_size_continuous(range = c(0.5, 0.5*max(log1p((data$size))))) +
  theme_classic() + guides(color=guide_legend(ncol=2))
p_cluster
```

### Differential expression analysis

We perform diffrential analysis to identify the markers of our cluster 3 as an example using the `FindMarkers` function. 
```{r r-mc-DEG, eval=TRUE}
# Set idents to metacell annotation 
Idents(MC.seurat) <- "SCclustering"

cells_markers <- FindMarkers(MC.seurat, ident.1 = "11", only.pos = TRUE)
cells_markers[order(cells_markers$avg_log2FC, decreasing = T)[1:10], ]
```

We see that the top marker genes for this cluster contain the killer cell lectin-like receptor (KLR) family, 
which is a group of transmembrane proteins preferentially expressed in NK cells. 

```{r r-mc-plot-genes, eval=TRUE, cache = FALSE}
genes = c("KLRF1", "KLRD1")
VlnPlot(MC.seurat, genes, ncol = 2, pt.size = 0.0)
```

We can verify the identification of the NK cell cluster by comparing the metacell annotation and the metacell clustering.

```{r fig.height=4,fig.width=10}
p_cluster + p_annot
```

### Visualize gene-gene correlation 

We can use the `supercell_GeneGenePlot` function from the SuperCell package to visualize the correlation between marker genes of a cell-type: 
(i) at the single-cell level and
(ii) at the metacell level.

For that, we load the single-cell data from which the metacells were derived from.
```{r r-load-single-cell, eval=TRUE}
print(proj_name)
sc_data <- readRDS(paste0("data/", proj_name, "/singlecell_seurat_filtered.rds"))
sc_data <- NormalizeData(sc_data, normalization.method = "LogNormalize")
```

We visualize gene-gene correlation at the single-cell level:
```{r r-sc-gene_gene_cor, eval=TRUE, fig.height=4,fig.width=8}
cells_markers <- cells_markers[order(cells_markers$avg_log2FC, decreasing = T),]
gene_x <- rownames(cells_markers)[1:5]  
gene_y <- rownames(cells_markers)[6:10]

alpha <- 0.7

p.sc <- SuperCell::supercell_GeneGenePlot(
  GetAssayData(sc_data, slot = "data"),
  gene_x = gene_x,
  gene_y = gene_y,
  clusters = sc_data@meta.data[, annotation_column],
  sort.by.corr = F, 
  alpha = alpha, 
  color.use = celltype_colors
)
p.sc$p
```

We visualize gene-gene correlation at the metacell level:
```{r r-mc-gene_gene_cors, eval=TRUE}

p.MC <- SuperCell::supercell_GeneGenePlot(GetAssayData(MC.seurat, slot = "data"),
                                          gene_x = gene_x,
                                          gene_y = gene_y,
                                          clusters = MC.seurat@meta.data[, annotation_column],
                                          sort.by.corr = F, supercell_size = MC.seurat$size,
                                          alpha = alpha,
                                          color.use = celltype_colors)
p.MC$p
```
